Confronting preferential sampling in wildlife surveys: diagnosis and model-based triage \footnotemark[2]

### 2016-06-24 manuscript/pref_sampling.tex
Restructured R files for package compilation and public consumption (including putting TMB .cpp code in the src directory for automatic compilation when package is loaded). Final changes to ms. before submission.

By contrast, CSPS (hereafter, PS), the primary focus of this article, arises when the locations being sampled and the process of interest (e.g. density, occupancy) are conditionally dependent given modeled covariates \citep{DiggleEtAl2010}.  For instance, PS can occur when the investigator uses a priori knowledge or observations of the state variable obtained during sampling to allocate survey effort in places where abundance or occurrence is known to be high. \citet{DiggleEtAl2010} showed that this type of PS can lead to bias when this extra information is not included in models for the state variable of interest.  Specifically, PS arises when we consider the set of sampled locations as stochastic and when $[{\bf R},{\bf Z} | {\bf x}] \neq [{\bf R}|{\bf x}][{\bf Z}|{\bf x}]$ \citep{DiggleEtAl2010}, where \textbf{R} is an indicator vector whose elements $R_i$ are 1.0 if sampling unit $i$ is sampled and are zero otherwise.  We use this definition of PS throughout the rest of the manuscript, noting that it is somewhat different than has sometimes been used in the SDM literature.  For instance, \citet{MerckxEtAl2011} use the term ``preferential sampling" to refer to the process of visiting some sites more often than others, while \citet{ManceurKuhn2014} define it as occurring when the locations selected for sampling are a function of an environmental covariate.  Neither of these latter conditions are problematic outside of the specialized field of presence-only modelling.

@@

Given these limitations, our present task is to generalize PS models to the types of data more typical of SDMs, and to allow the degree of PS to vary across the landscape.  Like \citet{DiggleEtAl2010} and \citet{PatiEtAl2011}, we impose a joint model for the process of interest (animal abundance or occurence) and the locations chosen for sampling. For the abundance process model, we start with eq. \ref{eq:process} as a general formulation for non-Gaussian data, writing the link-scale expectation as in eq. \ref{eq:mu}.
Next, recalling that $R_i$ is a binary indicator taking on the value 1.0 if survey unit $i$ is selected for sampling, and is 0.0 otherwise, we model $R_i$ using a Bernoulli distribution:
\begin{eqnarray}
 \label{eq:R}
  R_i & \sim & \text{Bernoulli}(h^{-1}(\nu_i)),
\end{eqnarray}
where $h()$ denotes a link function appropriate for binary data (e.g. logit, probit).  We then write the intensity for this model as
\begin{eqnarray}
  \label{eq:nu}
  \nu_i & = & \beta_0^* + {\bf x}_i^* \boldsymbol{\beta}^*+\eta_i+{\bf B} \delta_i.
\label{eq:nu}
\end{eqnarray}
 In a similar fashion to the model for the state process, the sampling intensity model has an intercept ($\beta_0^*$), explanatory covariates (${\bf x}_i^*$), fixed effect regression parameters ($\boldsymbol{\beta}^*$), spatially autocorrelated random effects ($\eta_i$ and $\delta_i$), and normally distributed error $\varepsilon_i$.  The predictive covariates ${\bf x}_i$ from Eq. \ref{eq:mu} and ${\bf x}_i^*$ from Eq. \ref{eq:nu} may or may not be the same.  Note also that the  spatially autocorrelated random effect $\delta_i$ is included in both Eqs. \ref{eq:mu} and \ref{eq:nu}, allowing for dependency in the two models, with the matrix {\bf B} describing the strength and type of dependence between the sampling process and underlying density.  The spatially autocorrelated random effects $\bfeta$ are assumed independent of the $\bfdelta$.  In practice, we find we often need to fix $\beta_0^*=0.0$ when random effects in Eq. \ref{eq:nu} are estimated to permit parameter identification.

@@

To illustrate PS and demonstrate that our proposed model has reasonable performance, we conducted a small simulation experiment.  For each of 500 simulations, we generated abundance of a hypothetical species over a $25 \times 25$ grid as
\begin{eqnarray*}
  [N_i|\mu_i] & = & \textrm{Poisson}(\exp(\mu_i)),
\end{eqnarray*}
where $i$ indexes survey unit $i$, and $\mu_i$ is determined according to Eq. \ref{eq:mu}. Abundance was generated as a function of a single spatially autocorrelated landscape covariate, as well as residual spatial autocorrelation ($\delta_i$) and overdispersion (fig. \ref{fig:PS_sims}). Specific details of of data generation procedures are presented in Appendix S1.

@@

Estimates of cumulative animal abundance across simulated landscapes were median unbiased for both estimation methods when the sites selected for sampling were independent of animal density, though when $b$ was estimated abundance estimates were more right skewed and had higher variance (fig. \ref{fig:sim_bias}).  Under moderate PS ($b=1$), estimation of the PS parameter $b$ led to a median bias of 5\%, while the canonical SDM model ignoring preferential sampling had a median bias of 40\%.  Under pathological PS ($b=5$), both estimation methods were extremely biased, but was even more severe for the naive model ignoring PS (fig. \ref{fig:sim_bias}).

@@

Integrated AIC strongly favored models with covariate effects, but for such models the presence of PS was equivocal (Table \ref{tab:aic}).  Further intuition can be gained by examining estimates of the PS parameter, $b$.  For the PS model without predictive covariates,  $\hat{b}=0.27$ (SE 0.11), and for the model with predictive covariates, $\hat{b}=0.19$ (SE 0.13).  Thus, it appeared that including predictive covariates decreased the PS effect size, as suggested by \citet{PatiEtAl2011}.  Estimates of abundance were substantially higher for models without a PS effect, with the non-PS model having a 49\% higher estimate when covariates were not modeled, and a 19\% higher estimate when covariate effects were included.

@@

Bias attributed to PS may seem counterintuitive, especially given the maxim in survey sampling to allocate more effort to strata for which animal density is high. For instance, in large scale line transect surveys under stratified sampling, the optimal amount of effort that should be allocated to stratum $s$ is $A_s D_s^{0.5}$, where $A_s$ is the area of $s$ and $D_s$ is the anticipated density \citep[][eqn 7.7]{BucklandEtAl2001}.  Thus, there are theoretical reasons to sample more in high density areas than in low density areas. The obvious solution in this instance is to compensate for PS in model-based inferences by accounting for variation in sampling intensity with explanatory covariates or post hoc stratification. However, it is not always clear how to perform such stratification when effort is allocated in a subjective manner.

When applied to bearded seal count data, approximately equal support was given to models with and without a PS effect.  The PS effect size was estimated to be positive and to produce considerably lower abundance estimates than models without a PS effect.  Differences between apparent abundance estimates decreased when covariates were added to model structure, supporting previous theoretical results \citep{DiggleEtAl2010,PatiEtAl2011} that covariates serve to decrease the conditional dependence between site selection and the state variable of interest.  However, in our data set, adding covariates did not eliminate evidence of PS. Accounting for PS in a model averaging framework led to a moderate decrease in our apparent abundance estimate for bearded seals in this region, and markedly decreased precision.  As in our simulations, the need to account for PS thus appeared to have a real cost in terms variance inflation.

>>>>

By contrast, CSPS (hereafter, PS), the primary focus of this article, arises when the locations being sampled and the process of interest (e.g. density, occupancy) are conditionally dependent given modeled covariates \citep{DiggleEtAl2010}.  For instance, PS can occur when the investigator uses a priori knowledge or observations of the state variable obtained during sampling to allocate survey effort in places where abundance or occurrence is known to be high. \citet{DiggleEtAl2010} showed that this type of PS can lead to bias when this extra information is not included in models for the state variable of interest.  Specifically, PS arises when we consider the set of sampled locations as stochastic and when $[{\bf R},{\bf Z} | {\bf x}] \neq [{\bf R}|{\bf x}][{\bf Z}|{\bf x}]$, where \textbf{R} is an indicator vector whose elements $R_i$ are 1.0 if sampling unit $i$ is sampled and are zero otherwise.  We use this definition of PS throughout the rest of the manuscript, noting that it is somewhat different than has sometimes been used in the SDM literature.  For instance, \citet{MerckxEtAl2011} use the term ``preferential sampling" to refer to the process of visiting some sites more often than others, while \citet{ManceurKuhn2014} define it as occurring when the locations selected for sampling are a function of an environmental covariate.  Neither of these latter conditions are problematic outside of the specialized field of presence-only modelling.

@@

Given these limitations, our present task is to generalize PS models to the types of data more typical of SDMs, and to allow the degree of PS to vary across the landscape.  Like \citet{DiggleEtAl2010} and \citet{PatiEtAl2011}, we impose a joint model for the process of interest (animal abundance or occurence) and the locations chosen for sampling. For the abundance process model, we start with eq. \ref{eq:process} as a general formulation for non-Gaussian data, writing the link-scale expectation as in eq. \ref{eq:mu}.
Next, recalling that $R_i$ is a binary indicator taking on the value 1.0 if survey unit $i$ is selected for sampling, and is 0.0 otherwise, we model $R_i$ using a Bernoulli distribution:
\begin{eqnarray}
 \label{eq:R}
  R_i & \sim & \text{Bernoulli}(h^{-1}(\nu_i)),
\end{eqnarray}
where $h()$ denotes a link function appropriate for binary data (e.g. logit, probit).  We then write the intensity for this model as
\begin{eqnarray}
  \label{eq:nu}
  \boldsymbol{\nu} & = & \beta_0^* + {\bf x}^* \boldsymbol{\beta}^*+\boldsymbol{\eta}+{\bf B} \boldsymbol{\delta}.
\label{eq:nu}
\end{eqnarray}
 In a similar fashion to the model for the state process, the sampling intensity model has an intercept ($\beta_0^*$), explanatory covariates (${\bf x}_i^*$), fixed effect regression parameters ($\boldsymbol{\beta}^*$) and spatially autocorrelated random effects ($\boldsymbol{\eta}$ and $\boldsymbol{\delta}$).  The predictive covariates ${\bf x}_i$ from Eq. \ref{eq:mu} and ${\bf x}_i^*$ from Eq. \ref{eq:nu} may or may not be the same.  Note also that the  spatially autocorrelated random effects $\boldsymbol{\delta}$ are included in both Eqs. \ref{eq:mu} and \ref{eq:nu}, allowing for dependency in the two models, with the matrix {\bf B} describing the strength and type of dependence between the sampling process and underlying density.  The spatially autocorrelated random effects $\bfeta$ are assumed independent of the $\bfdelta$.  In practice, we find we often need to fix $\beta_0^*=0.0$ when random effects in Eq. \ref{eq:nu} are estimated to permit parameter identification.

@@

To illustrate PS and demonstrate that our proposed model has reasonable performance, we conducted a small simulation experiment.  For each of 500 simulations, we generated abundance of a hypothetical species over a $25 \times 25$ grid as
\begin{eqnarray*}
  [N_i|\mu_i] & = & \textrm{Poisson}(\exp(\mu_i)),
\end{eqnarray*}
where $i$ indexes survey unit $i$, and $\mu_i$ is determined according to Eq. \ref{eq:mu}. Abundance was generated as a function of a single spatially autocorrelated landscape covariate, as well as residual spatial autocorrelation ($\delta_i$) and overdispersion (fig. \ref{fig:PS_sims}). Specific details of data generation procedures are provided in Appendix S1.

@@

Estimates of cumulative animal abundance across simulated landscapes were median unbiased for both estimation methods when the sites selected for sampling were independent of animal density, though when $b$ was estimated, abundance estimates were more right skewed and had higher variance (fig. \ref{fig:sim_bias}).  Under moderate PS ($b=1$), estimation of the PS parameter $b$ led to a median bias of 5\%, while the canonical SDM model ignoring preferential sampling had a median bias of 40\%.  Under pathological PS ($b=5$), both estimation methods were extremely biased, but was even more severe for the naive model ignoring PS (fig. \ref{fig:sim_bias}).

@@

Marginal AIC strongly favored models with covariate effects, but for such models the presence of PS was equivocal (Table \ref{tab:aic}).  Further intuition can be gained by examining estimates of the PS parameter, $b$.  For the PS model without predictive covariates,  $\hat{b}=0.27$ (SE 0.11), and for the model with predictive covariates, $\hat{b}=0.19$ (SE 0.13).  Thus, it appeared that including predictive covariates decreased the PS effect size, as suggested by \citet{PatiEtAl2011}.  Estimates of abundance were substantially higher for models without a PS effect, with the non-PS model having a 49\% higher estimate when covariates were not modeled, and a 19\% higher estimate when covariate effects were included.

@@

Bias attributed to PS may seem counterintuitive, especially given the maxim in survey sampling to allocate more effort to strata for which animal density is high. For instance, in large scale line transect surveys under stratified sampling, the optimal amount of effort that should be allocated to stratum $s$ is $A_s D_s^{0.5}$, where $A_s$ is the area of $s$ and $D_s$ is the anticipated density \citep[][eqn 7.7]{BucklandEtAl2001}.  Thus, there are theoretical reasons to sample more in high density areas than in low density areas. The obvious solution in this instance is to account for variation in sampling intensity with explanatory covariates or post hoc stratification. However, it is not always clear how to perform post hoc stratification when effort is allocated in a subjective manner.

When applied to bearded seal count data, approximately equal support was given to models with and without a PS effect.  The PS effect size was estimated to be positive and to produce considerably lower abundance estimates than models without a PS effect.  Differences between apparent abundance estimates decreased when covariates were added to model structure, supporting previous theoretical results \citep{DiggleEtAl2010,PatiEtAl2011} that covariates serve to decrease the conditional dependence between site selection and the state variable of interest.  However, in our data set, adding covariates did not eliminate evidence of PS. Accounting for PS in a model averaging framework led to a moderate decrease in our apparent abundance estimate for bearded seals in this region, and markedly decreased precision.  As in our simulations, the need to account for PS thus appeared to have a real cost in terms of variance inflation.
